### 2021-12-01 plos_latex_template.tex
RJ: Edited rule 5 and cleaned comments on rule 4

When choosing or inviting people for visible and valued roles in the conference––such as keynote speakers, program/scientific committee members, session chairs––, it is likely that there will not be much diversity in the first set of names.
Our biased lists are products of the existing systems that have always privileged some groups of people \cite{dwyerNoticeWhoScience2021,swartzScienceValueDiversity2019,wongBuildDiversityScience2020,dignazioUnicornsJanitorsNinjas2020}. 
Rather than deter us, this should encourage us to go beyond our narrow and often limited networks to look for, reach out to, invite, encourage, and onboard great people that are not routinely in the spotlight, making sure that these roles do not stay with people in privileged groups (e.g. avoid `manels' and all-white panels, for instance).
Make sure the composition of all these groups is balanced across topics, areas, and representation across regions, gender, race, career stage, among other aspects.
%rita: For academic conferences consider also to select for a talk master and PhD students and not to give voice only to senior scientist or principal investigator. lo pongo acá porque sí faltan cosas prácticas. % RJ: I added career stage to the sentence above. That should be enough.

@@

The final list of accepted talks can also be biased in both submission and acceptance processes. 
People who have never participated in conferences, and especially those regarded as big and prestigious, may not feel confident to submit their work.
To counteract this self-selection bias, promote the submission call beyond the communication channels and reach out to local groups and communities of practice to encourage submissions by their members.
Some of these communities may have pre-submission mechanisms to help their members prepare and improve their submissions (e.g. the Latin-R community has a Slack channel with the purpose of helping community members improve their abstracts for conferences, and the R-Ladies community did the same for useR! 2021).%sin el ejemplo no se entiende
% 1. Clear rules on reviews: double blind, double open, kind of open, etc.
% Be transparent on the evaluation criteria
% 2. Diverse program committee (above)
% 3. Unconscious bias (test)
% 4. Curate the final list against 
%see: https://www.science.org.au/files/userfiles/support/emcr/documents/one-page-summary-emcr-improving-diversity-web.pdf
% Unconscious (implicit) bias is the automatic impact of attitudes or stereotypes on our on our view of the world, actions, and decisions without our awareness (swartz 2019)
% in this study, we defined implicit bias as associations, attitudes, or beliefs that exist and enact their influence outside of an individual’s conscious awareness.
On the other hand, the review process may be biased because selection committees—even diverse ones—can be subjected to unconscious or implicit bias.
Unconscious bias happens when we interpret information, such as names of the authors, we assume their origin or ethnicity, their institution of origin, and we attach a positive or negative value to their work according to negative or positive stereotypes. Nobody is free from implicit bias [cite Ross2019].
Selection committees need to be reminded of such biases (the implicit association test can help identifying them [cite Greenwald - Measuring individual differences in implicit cognition: The implicit association test]), and
decide on the most appropriate strategy to counteract them, either with double blind, double open, or single blind reviews [cite discover cookbook].
Double blind reviews can help counteract implicit bias only if it is possible to completely anonymize the identifying information from submission. 
Double open, though more transparent, may still be subjected to implicit bias, and reviewers from minoritized groups might feel additional pressure because the reviews are signed. If this likely case, and in fields were it is not possible to guarantee anonymity of the authors during the selection process (e.g.  non-anonymized github repositories need to be examined), single blind reviews are the only option left.  
In any case, create clear evaluation criteria for the submitted proposals that do not put people from minoritized groups at disadvantage (e.g. based on their number of publications or talks), and share them with the authors and reviewers.
Advertising your practices to mitigate bias in proposal selection may also encourage people from diverse groups to submit their work, thus counteracting self selection bias.

@@

%Defy the stereotypical criteria for merit and how it is traditionally measured by acknowledging these community practices and the people behind them.
% You can also reframe the awards ceremony to acknowledge work beyond the traditional definition of merit/quality and include those who contributed to community building, presented teaching materials, and assumed other care-taking roles.

@@

% Rocío: main idea of the paragraph: online conferences can open opportunities for inclusion
%1. in 2021 an inclusive conference has to be an online or hybrid conference. this means a radical change in the functioning of all spaces, to include people who are online for any reason. we have elements to advocate for hybrid and lots of refs about online. cite joo and online refs. 
%practical: choose the platforms and be aware that reach varies between countries. check accessibility. understand bandwidth limitations. online does not make it automatically inclusive.
%2. hybrid. we advocate for active hybrid and distributed hybrid conferences because it's the most inclusive formats (not passive hybrid or semi-passive hybrid). we want all the inclusiveness of the in-person component with the _practical_ adaptations for hybrid (chat platform, q and a, streaming and networking) being online or in-person should not affect the possibility to be a chair a keynote or a presenter. no need to be present. a potential problem: opportunities of networking.
As stated in [the nature correspondence]: The COVID pandemic forced us to embrace online conferences (see C. Woolston Nature 582, 135–136; 2020), which remove some of the barriers that disproportionately affect marginalized groups. These include the cost of registration, transport and accommodation, the logistics of long-distance travel, and discriminatory visa applications (H. J. Niner \& S. N. Wassermann Front. Mar. Sci. 8, 638025; 2021; Gewin, V. Nature 569, 297–299; 2019). Inclusive conferences must have an online component [cite all the other references].
In-person interaction at conferences is priceless but often this is true and possible only for the ones who can afford to attend. 
Barriers such as cost of registration, transport and accommodation, the logistics of long-distance travel, and discriminatory visa applications, are particularly true for conferences that usually take place in high-income countries \cite{arendDisparityConferenceRegistration2019,gewinWhatScientistsShould2019,jooKeepOnlineOption2021}. 
Online conferences are more inclusive because they do not require a visa or a big budget, and are more accessible to people who may be unable to travel because of health issues or family responsibilities \cite{salibaGettingGripsOnline2020}.
This means that online conferences have a greater reach, not only in terms of attendees but in terms of speakers  \cite{atkinsonJournalMedicine20202021, roosOnlineConferencesNew2020, jooKeepOnlineOption2021}.
The online format may also make it easier to be inclusive of geographic regions by encompassing several time zones or deciding to rotate the favored time zone every year, without depending on a conference central location.

%Saranjeet: About time zones, although we can rotate and have different days dedicated to different time zones, we could also try spreading the conference time to be longer. So that each day the conference runs live for a couple of hours or so and the recorded material is made available/uploaded on YouTube (say) soon (every day). This might help because, not everyone might have the luxury of attending 4-6 hours of the conference on a weekday, even though it is running in their local timezone. However, people might be able to attend a couple of hours even on weekdays and will not get the feeling that they have missed a huge chunk of the conference. 
%ast add something about online giving flexibility in formats and how that depends on to the context of the conference
% Rocío: main idea: Networking may seem the weakest part of online conference, but it doesn't have to be. 
Networking and socializing have been mentioned as challenging aspects of online conferences, mostly because we are used to interactions at in-person social settings such as coffee or lunch breaks \cite{salibaGettingGripsOnline2020, roosOnlineConferencesNew2020}. 
%However, in-person settings might not be comfortable or appealing to everyone for socializing and some of these spaces may be exclusionary (e.g. galas or dinner nights at expensive venues). 
However, there is evidence that virtual communication before face-to-face interaction can make people from minoritized groups feel more included, thus participate more (e.g. \cite{trianaDoesOrderFacetoFace2012,blackEngenderingBelongingThoughtful2020}.
% JPNG: "On the other hand, there is evidence that virtual communication before face-to-face interaction can make people from minoritized groups feel more included, thus participate more (e.g. [29, 30]). Written media like chats create opportunities for long term interactions and networking by allowing more people to join the conversation even when it has started some time ago."
Organizers of online conferences should invest time in creating opportunities to meet and bond virtually, respecting people's limits, preferences, and remembering that `the usual' does not necessarily work for everyone, and that no single networking activity will ever serve the whole community. 
It is worth trying new and varied activities centered around different groups of people outside the mainstream of your conference.
%la: see https://www.trainingforchange.org/wp-content/uploads/2017/11/Mainstream-Margin-in-Groups.pdf
Some ideas in this line are: offering the option of written chat in addition to voice or video conversations, opening events with teamwork like trivia, offer some events that can be enjoyed passively like movies, yoga sessions, or art displays, where attendants can choose to just sit and enjoy without talking, or have a chat channel to comment on their experiences during the session.\fede{another activity: virtual city tours, I had it as a corollary activity for a summer school} \la{Some more specifics on how to achieve these suggestions in this paper: https://osf.io/k3bfn/}
% we need to check if something here (newbies!) is not missing in the above paragraphs
% As we mentioned previously, conference attendees should have ample opportunities to network with each other. However, having a diverse offer of networking opportunities that can appeal to people with different backgrounds, accessibility needs, and preferences can be challenging, especially if you have the mindset of organizing each activity for the complete pool of attendants. If you think of smaller activities that can reach specific groups, preferably co-led by community leaders, these sessions can be more productive successful, and inclusive than trying to organize one single activity that can please the whole community (\textbf{Rule \ref{rule_unbias}}). Some examples are the newbies sessions for first-timers, mixers lead by specific subgroups or communities, or leisure activities that reunite subgroups with the same interests: arts, exercise, sports, movies, etc.

% Rocío: main idea of the paragraph: In the context of reopening, hybrid conferences may be a thing. They also need to be inclusive.
Alternatively, a conference could have a hybrid format with an in-person and an online components that articulate well and maximize the experience of everyone attending your conference. This dual format could allow a group of people to interact face-to-face while providing many others the opportunity to participate remotely. 
Conference organizers who have been holding hybrid events since 2011 [Wenger-Tayner, 2021] consider that inclusive hybrid events are the future. These authors emphasize the need that all participants of hybrid events must go through the same event quality to the point that in-person and online participants pay the same registration amount to participate in high-quality hybrid conferences. These authors recommend having a buddy system, where every person attending online has an in-person attendant that makes sure the online person can follow the event and participate in the event as if they were in the room [Wenger-Trayner, 2020]. Wenger-Trayner (2020, 2021) also recommends having a chat function for back-channel conversations between all attendants, a virtual space for all online attendants to talk to each other, and allocating time to small group conversations, often mixing in-person and online groups. 
% Within the context of the COVID-19 pandemic, many conferences embraced the online format; but at the time of writing, some are reverting to in-person, which risks going back to the barriers mentioned above \cite{jooKeepOnlineOption2021}.
The challenge and requirement for this kind of setting would be to make the online component as relevant as the in-person component and not just a consolation prize to the less privileged in the community \cite{jooKeepOnlineOption2021,ninerBetterWhomLeveling2021}.
\rita{Sessions of an hybrid conference could have both chairs online, as attendee or speaker. This mean that is important when organizing the conference to include a third person who is in the venue to help the in-person participants, take the questions that are both online and in-person and communicate with the remote session chairs.}

\rita{The chat system of the conference can have a list of all the participants, with their registration name and if they are on-line or in-person, to facilitate the connections.}  The physical venue will have to include multiple ways to connect with online folks such as tablets on the meeting venue tables and cameras that allow online attendants to follow who is speaking in the in-person space. 
\rita{The venue selection must respond to some criteria such as give online support both bot in-person and online participants, to have stable and secure connection for the in-person participant to join the online event. Moreover, offer the option to create virtual room to connect in-person and online participant for networking events, such as scientific interest group, or LGBTQIA+-friendly spaces. Exhibition stands in the venue need to have facilities to connect with online participants (virtual stand). Furthermore, the poster session need to be carefully thought to give the best experience to both in-person and online participants. This should have both components to allow online participants to see the poster and have the possibility to ask questions. However, it also need allow who want to present their poster virtually to be seen by in-person participants and respond to questions about their works. Lastly to better plan your hybrid conference ask in advance who wish to travel to the venue and who wish to participate online. This will save you time and money for looking for a suitable venue.}

% [Wenger, 1999] https://doi.org/10.1017/CBO9780511803932
% [Wenger-Trayner, 2020] https://wenger-trayner.com/all/in-person-and-online-events/
% [Wenger-Trayner, 2021] https://wenger-trayner.com/reflections/in-person-online-hybrid-the-future/

>>>>

When choosing or inviting people for visible and valued roles in the conference––such as keynote speakers, program/scientific committee members, session chairs––, it is likely that there will not be much diversity in the first set of names.
Our biased lists are products of the existing systems that have always privileged some groups of people \cite{dwyerNoticeWhoScience2021,swartzScienceValueDiversity2019,wongBuildDiversityScience2020,dignazioUnicornsJanitorsNinjas2020}. 
Rather than deter us, this should encourage us to go beyond our narrow and often limited networks to look for, reach out to, invite, encourage, and onboard great people that are not routinely in the spotlight, making sure that these roles do not stay with people in privileged groups (e.g. avoid `manels' and all-white panels, for instance).
Make sure the composition of all these groups is balanced across topics, areas, and representation across regions, gender, race, career stage, among other aspects.

@@

The final list of accepted talks can also be biased in both submission and acceptance processes. 
People who have never participated in conferences, and especially those regarded as big and prestigious, may not feel confident to submit their work.
To counteract this self-selection bias, promote the submission call beyond the communication channels and reach out to local groups and communities of practice to encourage submissions by their members.
Some of these communities may have pre-submission mechanisms to help their members prepare and improve their submissions (e.g. the Latin-R community has a Slack channel with the purpose of helping community members improve their abstracts for conferences, and the R-Ladies community did the same for useR! 2021).
On the other hand, the review process may be biased because selection committees—even diverse ones—can be subjected to unconscious or implicit bias.
Unconscious bias happens when we interpret information, such as names of the authors, we assume their origin or ethnicity, their institution of origin, and we attach a positive or negative value to their work according to negative or positive stereotypes. Nobody is free from implicit bias [cite Ross2019].
Selection committees need to be reminded of such biases (the implicit association test can help identifying them [cite Greenwald - Measuring individual differences in implicit cognition: The implicit association test]), and
decide on the most appropriate strategy to counteract them, either with double blind, double open, or single blind reviews [cite discover cookbook].
Double blind reviews can help counteract implicit bias only if it is possible to completely anonymize the identifying information from submission. 
Double open, though more transparent, may still be subjected to implicit bias, and reviewers from minoritized groups might feel additional pressure because the reviews are signed. If this likely case, and in fields were it is not possible to guarantee anonymity of the authors during the selection process (e.g.  non-anonymized github repositories need to be examined), single blind reviews are the only option left.  
In any case, create clear evaluation criteria for the submitted proposals that do not put people from minoritized groups at disadvantage (e.g. based on their number of publications or talks), and share them with the authors and reviewers.
Advertising your practices to mitigate bias in proposal selection may also encourage people from diverse groups to submit their work, thus counteracting self selection bias.

@@

@@

% Rocío: main idea of the paragraph: online conferences can open opportunities for inclusion
Virtual conferences are more inclusive, removing barriers like costs of registration, transport, accommodation, the logistics of long-distance travel, and discriminatory visa applications \cite{jooKeepOnlineOption2021, ninerBetterWhomLeveling2021, salibaGettingGripsOnline2020, gichoraTenSimpleRules2010a}. 
After experiencing virtual conferences mostly due to the COVID pandemic, there is a widespread call to keep an online component in conferences, either maintaining conferences completely virtual or using hybrid formats combining in-person and online components \cite{jooKeepOnlineOption2021, woolstonLearningLoveVirtual2020, ninerBetterWhomLeveling2021, roosOnlineConferencesNew2020, levitisCenteringInclusivityDesign2021, sarabipourChangingScientificMeetings2021}.
Though more inclusive, an online component does not guarantee inclusion and accessibility for everyone.
Key decisions must be made early on in the planning stages \cite{levitisCenteringInclusivityDesign2021}. For instance, when choosing the conference platforms (for submission, registration, chatting system, live streaming, etc.) account for accessibility to screen readers, geopolitical restrictions, and bandwidth requirements. 
You should also decide how to be inclusive of geographic regions with the schedule of the conference, asking for pre-recorded videos, recording sessions, and sharing them during the conference. Some possibilities for scheduling are: having different days dedicated to different time zones; having each day with the same conference hours, and scheduling time slots that are friendly--and unfriendly--to different time zones; or having time-zone hubs in each day and repeating the events in each (in some cases live and others a recorded version).

% Rocío: main idea: Networking may seem the weakest part of online conference, but it doesn't have to be. 
Networking and socializing have been mentioned as challenging aspects of online conferences, mostly because we are used to interactions at in-person social settings such as coffee or lunch breaks \cite{salibaGettingGripsOnline2020, roosOnlineConferencesNew2020}. 
However, there is evidence that virtual communication before face-to-face interaction can make people from minoritized groups feel more included, thus participate more (e.g. \cite{trianaDoesOrderFacetoFace2012,blackEngenderingBelongingThoughtful2020}.
Invest time in creating opportunities to meet and bond virtually that can appeal to people with different backgrounds, disabilities, preferences, and remembering that no single networking activity will ever serve the whole community. 
Try new and varied activities centered around different groups of people outside the mainstream of your conference; if you think of smaller activities that can reach specific groups, preferably co-led by community leaders, these sessions can be more productive, successful, and inclusive.
Some examples are the newbies sessions for first-timers, social mixers lead by specific subgroups or communities like LGBT+, or leisure activities that reunite subgroups with the same interests: virtual art exhibitions, yoga sessions, movie watch parties, trivia quizzes, etc. 
Offer some activities that require voice or video interactions, and others that involve only written chat.

If you are organizing a hybrid event instead, you should decide on the relative importance that the online and in-person components will take. The most inclusive practice would be to give equal weight to both components, articulating them well and maximizing the experience of everyone attending your conference regardless of the format they choose to participate in. 
Under this format, the audience, chairs, and presenters--of talks, posters and workshops--could be either in-person or remote, and they should all be able to interact. 
For that purpose, all questions and comments should pass through an online system. 
The physical venue would have to include multiple ways to connect with online folks such as tablets on the meeting venue, cameras and microphones that allow online attendants to follow who is speaking in the in-person space.
This requires the venue to ensure stable and secure internet connection for the in-person participants to join the online folks. 
Having a hybrid format also means accounting for the fact that the venue does not need to host all the attendees. You should estimate in advance how many in-person attendees to expect, and ask them early on during registration or pre-registration. This could save you time and money when looking for a suitable venue.
When planning for networking activities in hybrid events, use the same principles of the paragraph above and offer varied activities centered around different groups. Some of these activities could be only in-person, but to promote connections between all participants, there should be a fair proportion of activities connecting in-person and online attendees.
